---
title: "Introduction to baytaAAR"
author: "Nils Müller-Scheeßel"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Introduction to baytaAAR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Load libraries

```{r, message=FALSE}
library(ggplot2)
library(baytaAAR)
library(dplyr)
library(posterior)
library(tidybayes)
```

## Introduction

**baytaAAR** implements latent trait analysis within a Bayesian Markov Chain Monte Carlo (MCMC) framework. It is intended to estimate the age-of-death of adult individuals for whom one or several ordinal traits have been assessed. It produces probability densities for the individual ages but also for the respective population as a whole. `baytaAAR` has been introduced and tested in the paper ###. More in-depth information on the background of and the rationale behind `baytaAAR` can be found in the vignettes.


## Data layout

As input, `baytaAAR` assumes a `matrix` of trait expressions. In its simplest form, this may contain only one column with a single trait as in the following example ("auricular surface").

```{r sorsum data, echo = F}
data(sorsum_as, package = "baytaAAR")
head(sorsum_as)
```

The example stems from the neolithic gallery grave Sorsum in Hessen/Germany. The auricular surface was the trait which was possible to assess most often [data from @ref_128246, table S69], and it was assessed according to the method by Lovejoy et al [-@ref_65832] which comprises eight levels. An overview of the distribution of the levels give the following plot:

```{r sorsum plot, echo = F}
ggplot(sorsum_as, aes(x =auricular_surface)) + geom_histogram(binwidth = 0.5) + theme_light() +
  scale_x_continuous(breaks=seq(1,8,1),minor_breaks=NULL) + xlab("\nauricular surface") + ylab("count\n")
```

Level 5 is the most numerous while levels 1 and 2 are only present once each.


## Running a first analysis

We will now run a first analysis with `baytaAAR`

```{r sorsum analysis, echo = TRUE}
set.seed(7281)
sorsum_as_res <- bay.ta.jags(
  method = as.matrix(sorsum_as[,2]),
  minimum_age = 18,
  parameters = c("b", "a", "beta0", "beta", "thresh", "age.s"),
  thinSteps = 1,
  numSavedSteps = 100000
)

sorsum_as_res_diag <- diagnostic.summary(sorsum_as_res,
                                         HDImass = 0.95, gelman_diag_multivariate = F)

# thresholds
#sorsum_as_res_diag[which(substring(rownames(sorsum_as_res_diag),
#                                   1, 11) == "thresh_age["),c(5,9,10)]

#sorsum_as_res_diag %>% .[-c(1:26),] -> sorsum_as_res_diag_age

# actual range of HDIs
#mean(sorsum_as_res_diag_age$HDIhigh - sorsum_as_res_diag_age$HDIlow)


#sorsum_as_res_diag_age %>% knitr::kable(., digits = 4, 
    #                                    caption = "sorsum. Diagnostic summary of age estimates.")

#sorsum_as_res_diag_age[-c(1:8, 11)] %>% knitr::kable(., digits = 1, 
#caption = "sorsum. Age estimate ranges (HDI) of a Bayesian model of unknown age.")

# plot to show age estimates
ggplot() + xlim(18, 100) + ylab("density") + 
  geom_function(fun = function(x) flexsurv::dgompertz(x - 18,
                                                      sorsum_as_res_diag[1,3], 
                                                      sorsum_as_res_diag[2,3])) +
    xlab("Gompertz distribution")

mcmc_array <- as_draws_array(sorsum_as_res)
draws_df <- as_draws_df(mcmc_array)

draws_df %>%
  dplyr::select("age.s[1]","age.s[2]","age.s[3]","age.s[4]",
                "age.s[5]","age.s[6]","age.s[7]","age.s[8]") %>%
  tidyr::pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") %>%
  ggplot(aes(x = value, y = parameter)) +
  stat_halfeye(.width = 0.95, point_interval = mode_hdi, fill = "lightgrey") +
  labs(
    x = "Modal age-at-death (years)",
    y = NULL
  ) + theme_light()

```


---

## References
