---
title: "Introduction to baytaAAR"
author: "Nils Müller-Scheeßel"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Introduction to baytaAAR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Load libraries

```{r, message=FALSE}
library(ggplot2)
library(baytaAAR)
library(dplyr)
library(posterior)
library(tidybayes)
library(tidyr)
```

## Introduction

**baytaAAR** implements latent trait analysis within a Bayesian Markov Chain Monte Carlo (MCMC) framework. It is intended to estimate the age-of-death of adult individuals for whom one or several ordinal traits have been assessed. It produces probability densities for the individual ages but also for the respective population as a whole. `baytaAAR` has been introduced and tested in the paper ###. More in-depth information on the background of and the rationale behind `baytaAAR` can be found in the vignettes. A further use-case discusses a real-life use-case with partly known age-at-death.


## Data layout

As input, `baytaAAR` assumes a `matrix` of trait expressions. In its simplest form, this may contain only one column with a single trait as in the following example ("auricular surface").

```{r sorsum data, echo = F}
data(sorsum_as, package = "baytaAAR")
head(sorsum_as)
```

The example stems from the neolithic gallery grave Sorsum in Hessen/Germany. The auricular surface was the trait which was possible to assess most often [data from @ref_128246, table S69], and it was assessed according to the method by Lovejoy et al [-@ref_65832] which comprises eight levels. An overview of the distribution of the levels give the following plot:

```{r sorsum plot, echo = F, fig.width=4, fig.height=3, fig.align = 'center' }
ggplot(sorsum_as, aes(x =auricular_surface)) + geom_histogram(binwidth = 0.5) + theme_light() +
  scale_x_continuous(breaks=seq(1,8,1),minor_breaks=NULL) + xlab("\nauricular surface") + ylab("count\n")
```

Level 5 is the most numerous while levels 1 and 2 are only present once each.


## Running a first analysis

We will now run a first analysis with `baytaAAR` with the Sorsum data. We will leave most parameters at the default values and only change the minimum age to `18` and the amount of saved steps to `1,000,000`. This number is divided by the number of chains (default = 3) to get at the total number of iterations, minus the number of iterations used for adaptation and burning-in.

```{r sorsum analysis, echo = TRUE}
sorsum_as_res <- bay.ta(
  method = as.matrix(sorsum_as[,2]),
  minimum_age = 18,
  numSavedSteps = 1000000
)
```

The analysis takes about 1 minute, depending on the computer power. The model output (`sorsum_as_res` in this case) is a matrix with the number of chains (three in the default case).

```{r sorsum diagnostics, echo = TRUE}
sorsum_as_res_diag <- diagnostic.summary(sorsum_as_res)
```

The output can be processed with any function able to deal with `coda` mcmc.lists but `baytaAAR provides some functions for convenience to quickly achieve results. One of them is the function `diagnostic.summary` which exactly does what the name suggests: a summary of diagnostic measures of the parameters.

```{r sorsum diagnostics table head, echo = FALSE}
sorsum_as_res_diag %>% head(., 20) %>% knitr::kable(., digits = 4)
```

A diagnostic table gives a first impression of the result. 

```{r sorsum diagnostic table, echo = T}
diagnostics.max.min(sorsum_as_res_diag)
```

Quality measures ...

```{r sorsum thresholds, echo = TRUE}
thresholds <- threshold.chains(sorsum_as_res)
thresh_diag <- diagnostic.summary(thresholds)
threshold.matrix(thresh_diag, mean_choice = "Mode") %>% data.frame(.) %>%
  knitr::kable(., digits = 4)
```

The thresholds of this traits.

```{r thresholds plot, echo = TRUE, warning=F, fig.width=5, fig.height=6, fig.align = 'center' }
mcmc_array <- posterior::as_draws_array(thresholds)
draws_df <- posterior::as_draws_df(mcmc_array)

draws_df[,c(1:7)] %>%
  tidyr::pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") %>%
  ggplot(aes(x = value, y = parameter)) +
  tidybayes::stat_halfeye(.width = 0.95, point_interval = mode_hdi, 
                          fill = "lightgrey") + 
  scale_x_continuous(breaks = seq(10,100,10), limits = c(15, 100)) +
  labs(
    x = "\nModal age-at-death (years)",
    y = NULL
  ) + theme_light() + theme(panel.grid.minor.x = element_blank())
```

Plots of the thresholds.

```{r sorsum agerange, echo = TRUE}
round(hdi.agerange(sorsum_as_res_diag), 1)
```

The age ranges

```{r sorsum ages, echo = TRUE, warning=F, fig.width=5, fig.height=6, fig.align = 'center' }
mcmc_array <- as_draws_array(sorsum_as_res)
draws_df <- as_draws_df(mcmc_array)
draws_df[,grep("^age.s\\[", colnames(draws_df))][,c(1:8)] %>%
  tidyr::pivot_longer(cols = everything(), names_to = "parameter", values_to = "value") %>%
  ggplot(aes(x = value, y = parameter)) +
  stat_halfeye(.width = 0.95, point_interval = mode_hdi, fill = "lightgrey") +
  scale_x_continuous(breaks = seq(10,100,10), limits = c(15, 100)) +
  labs(
    x = "\nModal age-at-death (years)",
    y = NULL
  ) + theme_light() + theme(panel.grid.minor.x = element_blank())
```

The plots of the first eight age estimates look similar to the threshold plots.

```{r sorsum Gompertz plot, echo = TRUE, fig.width=4, fig.height=3, fig.align = 'center' }
ggplot() + ylab("density\n") + 
  geom_function(fun = function(x) flexsurv::dgompertz(x - 18,
                                                      sorsum_as_res_diag[1,3], 
                                                      sorsum_as_res_diag[2,3])) +
    xlab("\nGompertz distribution") + theme_light() +
  scale_x_continuous(breaks = seq(10,100,10), limits = c(18, 100)) + theme(panel.grid.minor.x = element_blank())
```

Gompertz distribution.


## Going further

More vignettes are dedicated to ...

---

## References
