---
title: "Spitalfields: Comparison with known age-at-deah"
author: "Nils Müller-Scheeßel"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{known_age}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo = F}
library(baytaAAR)
library(dplyr)
library(ggplot2)
```

To show how the age estimations of `baytaAAR` can be compared with known age-at-death, we use the dataset of Spitalfields [@ref_263525]. The individuals come from a crypt at London ## and were identified in terms of age and sex by plates on their coffins.

```{r spitalfields data, echo = F}
data(spitalfields, package = "baytaAAR")
head(spitalfields)
```

Buckberry and Chamberlain introduced a new scoring system of the ## and thereby changed a single score with multiple levels to several traits with fewer levels each.

For running the model, with use NIMBLE and the multinorm likelihood (`mnorm`). `multicore` is set to `TRUE` to reduce computation time. Still, even on modern hardware, the model will run for several days until the usual quality criteria are met.

```{r spitalfields bayta, echo = TRUE}
spitalfields_res <- bay.ta(
  framework = "NIMBLE",
  algorithm = "mnorm",
  multicore = F,
  method = as.matrix(spitalfields[,c(2:6)]),
  minimum_age = 16,
  parameters = c("b", "a", "beta0", "beta", "thresh", "age.s", "Ustar"),
  thinSteps = 1,
  numSavedSteps = 10000,
  seed = 233
)
```

For once, we ignore the inspection of the diagnostics and go straight to the comparison with known age-at-death. For this, `bataAAR` provides several custom functions which essentially work as wrappers of other R-functions for convenience.

As the name suggests, the function `age.estimate.summarize` summarizes standard frequentist and Bayesian quality measures. It can make quite a difference which mean is chosen for the point estimate. These fall into three groups:

++

```{r spitalfields age.estimate.summarize, echo = TRUE}
age.comp.summarize(mcmc_list = spitalfields_res,
            known_age = spitalfields$Age,
            mean_choice = "Mode") %>% t(.) %>% round(.,2) %>%
  knitr::kable (., col.names = "value") %>%
  kableExtra::kable_styling(latex_options = c("HOLD_position"))
```

The age estimates are off

To test the deviation formally, in the literature the binomial test is usually invoked ##. The wrapper `sequential.binom.test` calculates this test for several given credibility levels.

```{r spitalfields binom test}
sequential.binom.test(spitalfields_res,
                      HDImass = c(seq(0.5, 0.9, 0.1), 0.95),
                      known_age = spitalfields$Age) %>%
  knitr::kable (.) %>%
  kableExtra::kable_styling(latex_options = c("HOLD_position"))
```

The result suggests that.

Finally, `age.comp.plot` shows the relationship between estimated and known age-at-death graphically. It can make quite a difference which mean is chosen for the point estimate.

```{r spitalfields plot, fig.width=8, fig.height=6, fig.align = 'center' }
spitalfields_diag <- diagnostic.summary(spitalfields_res)
age.comp.plot(spitalfields_diag, known_age = spitalfields$Age)
```

The plot demonstrate that

---

## References
