---
title: "Mathematical background"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Mathematical background}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Benjamin Gompertz and the Gompertz distribution

Central to `baytaAAR` is the Gompertz distribution, named after the British mathematican [B. Gompertz](https://en.wikipedia.org/wiki/Benjamin_Gompertz) (1779–1865) who first presented it in 1825 [@ref_110209]. 

Hazard, survival and the ensuing probability distribution are defined as follows [see also @wood_et_al_2002, 146]: 

$$
\begin{align}
\mu(t) & = \alpha \times \exp(\beta \times t) \\
S(t) & = \exp\big(\frac{\alpha}{\beta} \times (1-\exp(\beta \times t))) \\
f_0(t \mid \alpha,\beta) & = \alpha \times \exp\big(\beta \times t + \frac{\alpha}{\beta} \times (1-\exp(\beta \times t)))
\end{align}
$$

Despite or even because of its simplicity, it is still widely applied. Instrumental for `baytaAAR` is the strong empiric correlation between its two parameters $\alpha$ and $\beta$, the so-called *Strehler-Mildvan-correlation* [@strehler_mildvan_1960]. This makes it possible to reformulate it as essentially a one-parameter-distribution that still captures human mortality adequately. The empiric correlation between the two parameters at the age of 15 years was defined by T. Sasaki and O. Kondo [-@91048, 529 fig. 1] as follows:

$$
\begin{align}
Sab & = -2.624 \\
Sbb & = 0.0393 \\
Ma & = -7.119 \\
Mb & = 0.0718 \\
V & = 0.0823 \\
ln\alpha & \sim \mathrm{dnorm} (Sab \times (\beta - Mb)/Sbb + Ma, 1 / V ) \\
\alpha & = \exp(ln\alpha)
\end{align}
$$
Here, a normal distribution with a variance term of 0.0823 serves as basis to model logarithmically-scaled $\alpha$. If the variance term is omitted, the relationship between $\alpha$ and $\beta$ becomes deterministic:

$$
\alpha =\exp(Sab \times (\beta - Mb)/Sbb + Ma)
$$
If the starting age is not 15, we simulate the relation with the internal function `gomp.a0()`.

The parameter $\beta$ is estimated from the observed data, thus at the same time defining the overall level of mortality of the respective population and also narrowing down the age range of the individuals. The following figure, taken from Müller-Scheeßel et al. [-@ref_26972, 4 fig. 1], displays Gompertz hazard, survival and density curves with different $\beta$ values.

```{r gompertz_figure, warning=FALSE , fig.width=6, fig.height=5, include=TRUE, echo=FALSE, message=FALSE}
library(ggplot2)
library(gridExtra)
library(flexsurv)
# examples of Gompertz curves

# after Pflaumer 2011, 734
gomp_lx <- function(x, a, b) {
  lx <- exp(a/b - a/b * exp(b * x))
  return(lx)
}


# beta model values
beta1 <- 0.025
beta2 <- 0.04
beta3 <- 0.06
beta4 <- 0.09

# hgompertz(x, shape, rate): 
# x = age, shape = beta value, rate = derived from Sasaki & Kondo 2016 fig. 1, 2
# rate values according Sasaki & Kondo 2016 fig. 1, line 6, 30
Sab <- -2.624
Sbb <- 0.0393
Ma <- -7.119
Mb <- 0.0718
M1 <- Sab * (beta1 - Mb) / Sbb + Ma
M2 <- Sab * (beta2 - Mb) / Sbb + Ma
M3 <- Sab * (beta3 - Mb) / Sbb + Ma
M4 <- Sab * (beta4 - Mb) / Sbb + Ma

gridExtra::grid.arrange (
  
  ggplot()  + xlim(15, 100) + ylim(0, 0.4) +
    geom_function(fun = function(x) flexsurv::hgompertz(x - 15, 0.025, exp(M1)), 
                  aes(col = "\u03B2 = 0.025")) +
    geom_function(fun = function(x) flexsurv::hgompertz(x - 15, 0.04, exp(M2)),
                  aes(col = "\u03B2 = 0.04")) +
    geom_function(fun = function(x) flexsurv::hgompertz(x - 15, 0.06, exp(M3)), 
                  aes(col = "\u03B2 = 0.06")) +
    geom_function(fun = function(x) flexsurv::hgompertz(x - 15, 0.09, exp(M4)), 
                  aes(col = "\u03B2 = 0.9")) +
    ylab("hazard") + xlab("age in years") +
    theme_light() + 
    scale_colour_manual(values = c("red","blue","green", "dark grey")) +
    theme(legend.position = c(0.2, 0.7), legend.title = element_blank()),
  
  ggplot() + xlim(15, 105) +
    geom_function(fun = function(x) log(flexsurv::hgompertz(x - 15, 0.025, exp(M1))), 
                  colour = "red") +
    geom_function(fun = function(x) log(flexsurv::hgompertz(x - 15, 0.04, exp(M2))), 
                  colour= "blue") +
    geom_function(fun = function(x) log(flexsurv::hgompertz(x - 15, 0.06, exp(M3))), 
                  colour= "green") +
    geom_function(fun = function(x) log(flexsurv::hgompertz(x - 15, 0.09, exp(M4))), 
                  colour= "dark grey") +
    xlab("age in years") + ylab("hazard (log scale)") +
    theme_light(),
  
  ggplot() + xlim(15, 105) +
    geom_function(fun = function(x) flexsurv::dgompertz(x - 15, 0.025, exp(M1)), 
                  colour = "red") +
    geom_function(fun = function(x) flexsurv::dgompertz(x - 15, 0.04, exp(M2)), 
                  colour= "blue") +
    geom_function(fun = function(x) flexsurv::dgompertz(x - 15, 0.06, exp(M3)), 
                  colour= "green") +
    geom_function(fun = function(x) flexsurv::dgompertz(x - 15, 0.09, exp(M4)), 
                  colour= "dark grey") +
    xlab("age in years")  + ylab("density") +
    theme_light(),

  # gomp_lx() s. functions\helper_functions.R  
  ggplot() + xlim(15, 105) + ylim(0, 1) +
    geom_function(fun = function(x) gomp_lx(x - 15, exp(M1), 0.025), 
                  colour = "red") +
    geom_function(fun = function(x) gomp_lx(x - 15, exp(M2), 0.04), 
                  colour = "blue") +
    geom_function(fun = function(x) gomp_lx(x - 15, exp(M3), 0.06), 
                  colour = "green") +
    geom_function(fun = function(x) gomp_lx(x - 15, exp(M4), 0.09), 
                  colour = "dark grey") +
    ylab("survival") + xlab("age in years") +
    theme_light(),
  
  ncol = 2
)
```

The modal age _M_ is a useful parameter for adult mortality as it does not change with the starting observed age as life expectancy does [@missov_et_al_2015]. So, for example, e~20~ (life expectancy from age 20) will differ from e~25~ (life expectancy from age 25), and the one is not easily translated to the other. This is not the case with the modal age _M_. It equals the peak of the curves in the lower left image. The modal age _M_ is defined as:
$$
M = \frac{1}{\beta} \times log\big(\frac{\beta}{\alpha}\big)
$$


## The Gompertz distribution with NIMBLE and JAGS

The Gompertz distribution is implemented in NIMBLE and JAGS with the so-called "ones"-trick, which is a simple, convenient way to implement custom distributions [@ref_77051, 214--215]. Theoretically, NIMBLE allows to add custom functions as `nimble functions` compiled at runtime but we wanted to make sure that the results of the two frameworks are as similar as possible.


## Mathematical notation

The observed data are written as $y_{i,j}$ where \emph{i} indexes individuals from \emph{1 to N} and \emph{j} indexes traits (skeletal age “indicators”) from \emph{1 to J}. Because the data are recorded for ordinal categorical traits the values for each trait are in the form of positive integers. The number of states per trait is given as $K_j$. With $K_j$ states per trait $j$ there are $K_j - 1$ thresholds for each trait. 

$$
i = 1,\dots,N, \qquad
j = 1,\dots,J, \qquad
k = 1,\dots,K_j
$$

As stated above, the Gompertz model has two parameters $\alpha$ and $\beta$. The prior for $\beta$ is:

$$
\beta \sim \textit{U}(a = 0.02,\; b = 0.1)
$$
where \emph{U(a, b)} is a uniform distribution with the density function having a value of `1/(b - a)` between `a` and `b`. Simplifying the above equation of Sasaki and Kondo above, $\alpha$ is given deterministically as:

$$
\alpha = \exp(-66.77 \times \beta - 2.325)
$$

The prior for individual ages $t_i$, where $t$ is the age-at-death, is

$$
t_i \sim \textit{Gompertz}(\alpha,\; \beta)
$$
The starting age $t_0$ is usually at 15 years, and the distribution is truncated at `100 years - starting age`, reflecting a maximally expected life span of 100 years.

The thresholds within each trait are represented using the symbol $\gamma$, so that for a trait with five stages, the thresholds are $\gamma_{k_1 = 1} < \gamma_{k_1 = 2} < \gamma_{k_1 = 3} < \gamma_{k_1 = 4}$. So that the model is identifiable, all $\gamma_{k_j = 1} = 1.5$.

The priors within each trait $j$ for the thresholds with more than two states are:

$$
\gamma_{k=2,…,K_j} \sim \textit{TN}(
\mu = k + 0.5,\;
\sigma = 10,\;
a,\;
b = \infty)
$$

where \emph{TN()} is a truncated normal distribution with left truncation at `a` and right truncation at `b`. Because the thresholds must be ordered, for the JAGS-model, $a = 1.5$, and the simulated thresholds are sorted in ascending order. NIMBLE does not have a sort-function, so there $a = \gamma_{k-1}$.

For each individual there is a vector of latent traits $z_i$ of length $J$ (number of traits). For the simple probit model with conditional independence, the elements of these vectors are modeled separately with normal distributions:

$$
z_{i,j} \sim \mathcal{N}(\mu_{i,j},\; \sigma = 1)
$$

For identifiability, $\sigma$ is fixed at \emph{unity} (1). 

For the complex probit model, these vectors have a multivariate normal distribution:

$$
\boldsymbol{z}_i \sim \mathcal{MVN}(
{\mu}_{i,j},\; 
\mathbf{R}),
$$

where there is a “flat” prior by Lewandowski, Kurowicka, & Joe [-@ref_54568] for the correlation matrix:

$$
\mathbf{R} \sim \textit{LKJ}(\eta = 1)
$$

For $\eta = 1$, this prior is uniform over the space of correlation matrices.

The terms in the vector $\mu_{i}$ are:

$$
\mu_{i,j} = \textit{beta}_{0_j} + \textit{beta}_j \times \ln(t_i)
$$

with the priors for the $beta_0$ and $beta$ parameters within each trait $j$ being:

$$
\textit{beta}_{0_j} \sim \textit{TN}(\mu = 0,\; \sigma = 10,\; 
a = -\infty,\; b = 0)
$$

$$
\textit{beta}_{j} \sim \textit{TN}(\mu = 0,\; \sigma = 10,\; 
a = 0,\; b = \infty)
$$

JAGS and NIMBLE both have the distribution `dinterval` which is difficult to write in a succinct format:

$$
y_{i,j} =
\begin{cases}
1, & \text{if } z_{i,j} \le 1.5, \\[6pt]
k_j, & \text{if } \gamma_{j,k_j-1} < z_{i,j} \le \gamma_{j,k_j},
\quad k_j=2,\dots,K_j-1, \\[6pt]
K_j, & \text{if } \gamma_{j,K_j-1} < z_{i,j} 
\end{cases}
$$

Because `dinterval` indexes categories from `zero` instead of `one`, the observed trait levels and thresholds are internally shifted by one unit.

The proper initialisation of the chains is of eminent importance in probit regression. Each chain starts with different parameter values to enhance mixing, and most parameters pick random values from uniform distributions. For Gompertz $\beta$ the init value is in the range `0.02–0.1`, for $beta$ the range is `0.5–1`, and for $beta_0$ `–10––3`. The init value for age $t_i$ is `20–40`, assuming that the starting age is `0`. The correlation matrix is for all chains initialised with an identity matrix (ones in the diagonal and zeros otherwise). The thresholds  are initialized with `k + 0.5` and the vector of latent traits $z_i$ with $y_{i,j} - \textit{U}(-0.2, 0.2)$.

---

## References
