---
title: "Parallel_processing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parallel_processing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(baytaAAR)
```


# Parallel processing and number of chains

On modern hardware with usually more than one core, parallel processing is very useful as it cuts down time significantly. In a normal case, you would use as many cores as you specify chains. The default for JAGS and nimble is three chains. More chains can help in reaching quality criteria like psrf faster. However, it also increases the size of the MCMC samples so that post-processing in the end becomes a question of available memory. Therefore, and following Kruschke, we control the number of saved runs by dividing numsavedruns by the number of chains. So, increasing the number of chains will automatically shorten the number of saved runs which will usually lead to poorer mixing. In this respect, we made the observation for our model that a higher number of chains does not outweigh the disadvantage of shorter chains. Therefore, we actually do not advise to increase the number of chains.


## Parallel processing in JAGS

Parallel processing in JAGS is straightforward. In bay.ta.jags, you specify the mode of operation with the parameter "method". "rjags" equals single-core-processing, and with "rjparallel" and "parallel", you invoke parallel processing. "parallel" is the usual choice as it gives feedback during processing.


## Parallel processing in nimble

Parallel processing in nimble is much more complicated. The handbook gives a general account of the necessary steps (https://r-nimble.org/examples/parallelizing_NIMBLE.html). For a package like ours, there is the additional difficulty that we provide custom functions which have to be supplied in the environment of each cluster. However, simply copying the functions to the global environment would be unacceptable behaviour of a package.

In the end, we hope that we succeeded to make parallel processing with nimble as unintrusive as possible by providing a wrapper function "bay.ta.nimble" which takes care of the different modes of operation. Please note, though, that you will get no feedback during running the function in parallel. Therefore, it is good practice to start with a low number of steps and see how long these take, interpolate the number of necessary steps from the quality measures (especially ESS) and change the settings accordingly.

---
