---
title: "Chelsea 'Old Church': A worked example"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Chelsea 'Old Church': A worked example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Load libraries

```{r setup, message = F}
library(baytaAAR)
library(dplyr)
library(tidyr)
library(ggplot2)
```

## The churchyard of All Saints, Chelsea's Old Church

### Introduction

For the worked example to show the complete range of functions provided by `baytaAAR`, we chose the churchyard of All Saints, Chelsea's Old Church [@ref_29376]. For an overview of the site and the osteological investigations, see the respective website of the [Museum of London](https://www.londonmuseum.org.uk/collections/research/osteological-database/post-medieval-cemeteries/all-saints-chelsea-old-church-churchyard/).

The data of the osteological team of the Museum of London is freely available, including the age estimations and especially the raw scores that formed the basis for the finally published age ranges. If possible, the team applies four established age estimation methods to the human remains: Dental wear with Codes 1-4, Pubic symphysis with Phases I-VI, Auricular surface with Phases 1-8, and Costochondral with Phases 0-8. For the background of these methods and their application, see again the respective website of the [Museum of London](https://www.londonmuseum.org.uk/collections/research/osteological-database/bioarchaeology/). From the complete Chelsea dataset, we chose those individuals where at least one of the four methods was applicable. As a special feature of this churchyard, for some of these individuals the age-at-death is known thanks to plates on the coffins.

This worked example proceeds in five steps: (1) data preparation, (2) single-trait models, (3) multivariate models, (4) model comparison and calibration, and (5) evaluation against known ages. One of the main aims is to sketch the process to arrive at plausible age estimates in terms of *highest density intervals* when the data set contains a lot of missing instances. The complete-case model best captures joint trait covariance without imputation, hence provides the least biased estimate of the underlying uncertainty structure.


### Loading data

The first step is the loading and preparation of the data. For that, we download and extract the data directly from the website of the Museum of London and extract it. The second table contains known age-at-death for some of the individuals from plates on the coffins. We then filter only the adults and expand the age estimation table. The two tables are merged and only the `CONTEXT` column with the individual number, the sex and age estimation as well as the raw age estimations code and the known age-at-death column are retained. Finally, all entries `9`, which for all four age estimations methods equal `undecided`, are converted to `NA` and only those rows with at least one valid method are kept.

```{r loading of data, echo=T, warning=FALSE}
temp <- tempfile()
download.file(
  "https://dams.londonmuseum.org.uk/m/2dc89b01e62759ff/original/PMCOC.zip",
  temp, mode = "wb")
unzip(temp, "PMCOC_age_estimates.lst")
chelsea_age <- read.delim("PMCOC_age_estimates.lst",
                          sep = "|",
                          header = TRUE,
                          stringsAsFactors = FALSE,
                          row.names = NULL,
                          fill = TRUE,
                          strip.white = TRUE)

# absolute ages according to Cowie et al. 2008, 41 table 6
chelsea_known_age <- data.frame(
  CONTEXT = c(35, 147, 198, 462, 525, 622, 654, 681, 701, 713,
              722, 744, 792, 976, 980, 990, 1133),
  known_age = c(60, 67, 44, 61, 70, 84, 66, 84, 78, 68, 56, 70,
                68, 82, 54, 32, 70) )

chelsea_age_col <- colnames(chelsea_age)[-7]
chelsea_age <- chelsea_age[,-14]
colnames(chelsea_age) <- chelsea_age_col

chelsea_adults <- 
  chelsea_age |> dplyr::filter(grepl("^(ADULT|UNCLASSIFIED ADULT)", AGE))

chelsea_adults_wide <- chelsea_adults |>
  dplyr::mutate(VALUE = as.numeric(VALUE)) |>
  tidyr::pivot_wider(
    id_cols = c(CEMETERY, SITECODE, PERIOD, LU_INT, E_DATE, L_DATE,
                CONTEXT, SEX, AGE, TRAIT_TYPE),
    names_from = EXPANSION,
    values_from = VALUE
  )

# combine both tables
chelsea <- merge(chelsea_adults_wide, chelsea_known_age,
                     by = "CONTEXT",
                     all.x = TRUE)

# limit data set to relevant columns
chelsea_mat <- chelsea[,c(1, 8, 9, 12,13,14,15,18)]

chelsea_mat[chelsea_mat == 9] <- NA
chelsea_complete <- chelsea_mat[rowSums(is.na(chelsea_mat[,4:7])) != 4, ]
chelsea_complete$AGE <- factor(chelsea_complete$AGE,
                               levels = c("ADULT 18-25 YEARS",
                                          "ADULT 26-35 YEARS",
                                          "ADULT 36-45 YEARS",
                                          "ADULT >46 YEARS",
                                          "UNCLASSIFIED ADULT"),
                               ordered = TRUE)
chelsea_complete$SEX <- factor(chelsea_complete$SEX,
                          levels = c("FEMALE",
                                     "FEMALE?",
                                     "INTERMEDIATE",
                                     "MALE?",
                                     "MALE"),
                          ordered = TRUE)

# define data sets with specific traits
chelsea_no_nas <- chelsea_complete[rowSums(is.na(
  chelsea_complete[,4:7])) == 0, ]
chelsea_ps_as <- 
  chelsea_complete[rowSums(is.na(chelsea_complete[,5:6])) == 0, ]
chelsea_dental <- chelsea_complete[(!is.na(chelsea_complete[,4])), ]
chelsea_ps <- chelsea_complete[(!is.na(chelsea_complete[,5])), ]
chelsea_as <- chelsea_complete[(!is.na(chelsea_complete[,6])), ]
chelsea_costo <- chelsea_complete[(!is.na(chelsea_complete[,7])), ]

# show first lines, replace underscore with space for variable "known_age"
head(chelsea_complete) |> 
  knitr::kable (col.names = gsub("[_]", " ", names(chelsea_complete)))
```

The table above gives an impression of the data structure of the Chelsea individuals with estimated sex and age, the four traits used for estimations as well as a last column with known age-at-death for a few individuals.


### Descriptive statistics

Next, we want to get an overview of the number of individuals, the sex estimations and especially the number of phases for each method.

```{r number of individuals, echo=TRUE}
chelsea_complete |> nrow()
```

All in all, there are at least 152 individuals with information on at least one method.

```{r sex estimations, echo=TRUE}
chelsea_complete[,2] |> table()
```

Based on the osteological sex estimations, it seems that both sexes are present in roughly equal numbers. The known-age structure is as follows:

```{r age estimations, echo=TRUE}
chelsea_complete$known_age |> na.omit() |> as.numeric() |> sort()
```

 Most of the 16 individuals with known age-at-death are at least 60 years or above. There are only four individuals with a plate of younger age.
 
 The following code produces an overview of the age estimations methods and the counts per phase of each method, including the number of `NAs`:

```{r method overview, echo=TRUE}
options(knitr.kable.NA = '')
chelsea_complete |>
  dplyr::select(starts_with("Dental"),
         starts_with("Pubic"),
         starts_with("Auricular"),
         starts_with("Costo")) |>
  tidyr::pivot_longer(everything(), names_to = "method", values_to = "phase") |>
  dplyr::mutate(phase = as.character(phase)) |>
  dplyr::count(method, phase) |>
  tidyr::pivot_wider(names_from = phase, values_from = n, values_fill = NA) |>
  knitr::kable ()
```

The overview shows that the costochondral is the method with the least applicability (= most `NAs`) while the method with the least `NAs` is the auricular surface, followed by the pubic symphysis. Dental wear was possible to determine only in a few cases more than the costochondral. All in all, the missing information is considerable:

```{r numbers overview, echo=TRUE}
count_na <- sum(is.na(chelsea_complete[,4:7]))
count_cells <- nrow(chelsea_complete) * 4
round(count_na / count_cells * 100,1)
```

Overall, more than 40 percent of the cells are `NA`. This level of missingness is expected to affect multivariate models via imputation uncertainty.


## Running `baytaAAR`

We will now analyse the Chelsea data with `baytaAAR` systematically. This implies to analyse all methods separately and then together. We will do the following runs, also indicate if we use the `JAGS` or `NIMBLE` version as well as either `norm` (multiple normal) or `mnorm` (multinormal) in case of the inclusion of more than one method in the same run. The difficulty here is that many entries, as shown above, are `NA`. These will be a challenge for the multinormal analysis as the unobserved nodes will have to be imputed. We will compare a minimum set of individuals where all four methods are present with a maximum set of the two most common methods, auricular surface and pubic symphysis.

* only auricular surface (with `JAGS`)
* only costochondral (with `JAGS`)
* only dental wear (with `JAGS`)
* only pubic symphysis (with `JAGS`)
* all methods and only individuals with all methods present (with `NIMBLE` and `mnorm`)
* only auricular surface and pubic symphysis and only individuals with both methods present (with `NIMBLE` and `mnorm`)
* all methods and all individuals (with `NIMBLE` and `mnorm`)
* all methods and all individuals with calibration term (with `JAGS`)


### Defining paths for downloading RDA-files

Please note that the models should run until the resulting Markov Chain Monte Carlo (MCMC) samples meet certain quality criteria. As already noted in the introductory article (`vignette("baytaAAR")`), these are `10,000` for `ESS` and below `1.1` for `PSRF`. To spare the time in running the models, we saved RDA-files with the raw MCMC-chains on `Github` for downloading and loading. If you really want to run the models, set `runNewMCMC` to `TRUE`. If not, stick to the negation `!TRUE`.


```{r paths for rda-files}
runNewMCMC = !TRUE
# path <- "https://github.com/ISAAKiel/Chelsea_mcmc/raw/refs/heads/main/"
path <- "https://raw.githubusercontent.com/ISAAKiel/Chelsea_mcmc/main/"
```


### Only single traits (with `JAGS`)

We first concentrate on only those individuals where the auricular surface was assessable. The first argument `method` is the trait which are used for age estimation, the auricular surface in this case. The data has to be of the class `matrix`. `minimum_age` and `maximum_age` truncate the underlying Gompertz distribution to a reasonable range. `parameters` denotes the estimated and monitored variables. They have the following meaning:

* b: Gompertz beta
* a: Gompertz $\alpha$, in `bay.ta` linked deterministically to Gompertz $\beta$ to reduce volatility
* beta0: the intercept of the latent continuous variable underlying the ordinal trait
* beta: the slope of the latent continuous variable underlying the ordinal trait
* thresh: the threshold value. The number of thresholds equals the number of levels of the trait minus 1
* age.s: the estimated age of each individual

The rest of the variables control `JAGS`. `multicore` specifies here that the code is executed in parallel for speed. `thinSteps` controls the amount of thinning of the resulting Markov Chain. With `1` there is no thinning. However, because the ordered probit model suffers from high auto-correlation, this number will almost certainly not suffice. Setting the number of `numSavedSteps` too high might lead to overly large files. Therefore, it is always a good idea to start with a moderate value for `numSavedSteps` of about `10,000` and a value for `thinSteps` of 1, and then increase the number of steps according to the first result. A future version of `baytaAAR` might also include the possibility to define quality goals which the code should reach. Finally, we run `bay.ta` with a seed for reproducibility

```{r auricular surface with JAGS}
if ( runNewMCMC ) {
  chelsea_as_res <- bay.ta(
    framework = "JAGS",
    method = chelsea_as[,6],
    minimum_age = 18,
    maximum_age = 100,
    parameters = c( "b", "a", "beta0", "beta", "thresh", "age.s"),
    multicore = TRUE,
    thinSteps = 10000, # necessary 10000, takes 16 hours
    numSavedSteps = 80000,
    seed = 3901)
} else {
  file <- "chelsea_as_res.Rdata"
  temp <- tempfile()
  download.file(paste0(path, file), destfile = temp, mode = "wb")
  con <- gzfile(temp, "rb")
  load(con)
  close(con)
  threshold_as <- chelsea_as_res |> threshold.chains() |> 
  diagnostic.summary() |> threshold.matrix(mean_choice = "Mode")
}
```

The resulting Markov chains are stored in the file `chelsea_as_res` in the `coda::mcmc.list()` format for further processing. The function `diagnostic.summary()` provides a summary of the posterior estimates:

```{r auricular surface diagnostic summary}
chelsea_as_diag <- diagnostic.summary(chelsea_as_res)
head(chelsea_as_diag[-c(9:12),], 20) |>  knitr::kable (digits = 4)
```

The above table just shows a part of the full diagnostic table. Especially when many parameters are to be estimated, it can contain quite a number of rows. The columns show a summary of the individual estimates of each parameter with the following meaning:

- PSRF Point est.: *Potential scale reduction factor* (= Gelman-Rubin statistic), a measure of the mixing of chains. 
- PSRF Upper C.I.: The upper limit of the 0.95-confidence interval of the PSRF.
- Mean: Arithmetic mean of the estimates.
- Median: Median of the estimates.
- Mode: Mode of the estimates.
- ESS: *Effective sample size*, a control of autocorrelation.
- MCSE: *Monte Carlo standard error*.
- HDImass: Credibility level of the *highest density interval*
- HDIlow: Start of the *highest density interval*.
- HDIhigh: End of the *highest density interval*.

For more information on these measures, the reader is referred to e.g. the book "Doing Bayesian data analysis" by J. Kruschke [-@ref_77051]. The summary tool `diagnostics.max.min` quickly allows to see the worst values:

```{r auricular surface diagnostics max min}
chelsea_as_diag |> diagnostics.max.min()
```

The highest `PSRF` value is clearly far below `1.1` and the lowest `ESS` value higher than `10,000`. We are therefore safe to proceed with the analysis and to calculate a summary of the age estimates:

```{r auricular surface median age ranges}
chelsea_as_age_ranges <- age.estim.summary(chelsea_as_diag)
chelsea_as_age_ranges |>  knitr::kable(digits = 3)
```

From the diagnostics, the `baytaAAR` helper function `age.estim.summary()` computes, for example, that the average mean, median and mode of the age range estimated from the data is hovering around 45 years. This might seem like a very large range (and it certainly is!) but the model promises that this credible range contains 95% of the true ages-at-death.

We will now proceed with the other traits and look again at the auricular surface when comparing its results with the other runs. The models for the other single traits are not discussed in detail as they only mirror the approach for the auricular surface. We will examine their results when all model runs have been completed.


```{r other single method models, echo = TRUE}
if ( runNewMCMC ) {
  chelsea_ps_res <- bay.ta(
    framework = "JAGS",
    method = chelsea_ps[,5],
    minimum_age = 18,
    maximum_age = 100,
    parameters = c( "b", "a", "beta0", "beta", "thresh", "age.s"),
    multicore = TRUE,
    thinSteps = 2000,
    numSavedSteps = 100000,
    seed = 991)
} else {
  file <- "chelsea_ps_res.Rdata"
  temp <- tempfile()
  download.file(paste0(path, file), destfile = temp, mode = "wb")
  con <- gzfile(temp, "rb")
  load(con)
  close(con)
}
chelsea_ps_diag <- diagnostic.summary(chelsea_ps_res)
chelsea_ps_age_ranges <- age.estim.summary(chelsea_ps_diag)
threshold_ps <- chelsea_ps_res |> threshold.chains() |> 
  diagnostic.summary() |> threshold.matrix(mean_choice = "Mode")

if ( runNewMCMC ) {
  chelsea_costo_res <- bay.ta(
    framework = "JAGS",
    method = chelsea_costo[,7],
    minimum_age = 18,
    maximum_age = 100,
    parameters = c( "b", "a", "beta0", "beta","thresh","age.s"),
    multicore = TRUE,
    thinSteps = 1500,
    numSavedSteps = 100000,
    seed = 6433)
} else {
  file <- "chelsea_costo_res.Rdata"
  temp <- tempfile()
  download.file(paste0(path, file), destfile = temp, mode = "wb")
  con <- gzfile(temp, "rb")
  load(con)
  close(con)
}
chelsea_costo_diag <- diagnostic.summary(chelsea_costo_res)
chelsea_costo_age_ranges <- age.estim.summary(chelsea_costo_diag)
threshold_costo <- chelsea_costo_res |> threshold.chains() |> 
  diagnostic.summary() |> threshold.matrix(mean_choice = "Mode")

if ( runNewMCMC ) {
  chelsea_dental_res <- bay.ta(
    framework = "JAGS",
    method = chelsea_dental[,4],
    minimum_age = 18,
    maximum_age = 100,
    parameters = c( "b", "a",  "beta0", "beta", "thresh",  "age.s"),
    multicore = TRUE,
    thinSteps = 1000,
    numSavedSteps = 100000,
    seed = 5783)
} else {
  file <- "chelsea_dental_res.Rdata"
  temp <- tempfile()
  download.file(paste0(path, file), destfile = temp, mode = "wb")
  con <- gzfile(temp, "rb")
  load(con)
  close(con)
}
chelsea_dental_diag <- diagnostic.summary(chelsea_dental_res)
chelsea_dental_age_ranges <- age.estim.summary(chelsea_dental_diag)
threshold_dental <- chelsea_dental_res |> threshold.chains() |> 
  diagnostic.summary() |> threshold.matrix(mean_choice = "Mode")
```


### Only auricular surface and pubic symphysis, with correlation (`NIMBLE`)

Up to this point, traits were modeled independently. This ignores biological correlation between ageing indicators. The following models explicitly estimate this dependence via a multivariate normal structure. With more than one trait, correlation between the traits should be taken into account, and we now move on to a model with multivariate normal likelihood function. Our first model with conditional dependence is based on the individuals where both the auricular surfance and the pubic symphysis were assessed. This is the case for 83 individuals:

```{r only auricular surface and the pubic symphysis nimble model, n, echo = T}
  chelsea_ps_as |> nrow()
```

The multinormal function looks very similar to the multiple normal one. The only major difference is the specification of the algorithm as `mnorm` for "multinormal". Additionally, among the parameters, we specified `Ustar` which denotes the cholesky decomposition of the correlation matrix. The Lewandowski-Kurowicka-Joe distribution that functions as prior for the correlation matrix is controlled by the parameter `eta`. A value of `1` resembles an uninformative prior assuming no correlation between the traits.

```{r only auricular surface and the pubic symphysis nimble model}
if ( runNewMCMC ) {
  chelsea_ps_as_res <- bay.ta(
    framework = "NIMBLE",
    method = chelsea_ps_as[5:6],
    algorithm = "mnorm",
    eta = 1,
    multicore = TRUE,
    nChains = 3,
    minimum_age = 18,
    maximum_age = 100,
    parameters = c( "b", "a","beta0", "beta","thresh", "age.s", "Ustar"),
    thinSteps = 25000,
    numSteps = 50000,
    seed = 8992)
} else {
  file <- "chelsea_ps_as_res.Rdata"
  temp <- tempfile()
  download.file(paste0(path, file), destfile = temp, mode = "wb")
  con <- gzfile(temp, "rb")
  load(con)
  close(con)
  chelsea_ps_as_res <- coda::as.mcmc.list(chelsea_ps_as_res)
  chelsea_ps_as_res_diag <- diagnostic.summary(chelsea_ps_as_res)
  threshold_ps_as <- chelsea_ps_as_res |> threshold.chains() |> 
    diagnostic.summary() |> threshold.matrix(mean_choice = "Mode")
  chelsea_ps_as_age_ranges <- chelsea_ps_as_res_diag |> age.estim.summary()
}
```

As a shortcut, only the result of the `diagnostics.max.min()` is shown in the following table:

```{r auricular surface + pubic symphysis diagnostics max min}
chelsea_ps_as_res_diag |> diagnostics.max.min()
```

The quality measures seem to be satisfied, so we can now have a look at the correlation extracted with the helper function `corr.mat.mean()`.

```{r only auricular surface and the pubic symphysis nimble model, correlation}
corr.mat.mean(chelsea_ps_as_res)[1,2] |> round(3)
```

The result implies that the correlation between the auricular surface and the pubic symphysis is very low and below 0.1.

  
### All traits, with correlations (`NIMBLE`)
  
If we want to analyse all those individuale where four traits auricular surface, the pubic symphysis, the costochondral as well as dental wear were observed, we severely limit the available sample:
  
```{r all methods nimble model, no NAs, n, echo = T}
  chelsea_no_nas |> nrow()
```

Only 21 individuals fulfill the criterion.

```{r all methods nimble model, no NAs}
if ( runNewMCMC ) {
  chelsea_no_nas_res <- bay.ta(
    method = chelsea_no_nas[4:7],
    algorithm = "mnorm",
    eta = 1,
    multicore = TRUE,
    nChains = 3,
    minimum_age = 18,
    maximum_age = 100,
    parameters = c( "b", "a", "Ustar", "beta0", "beta", "thresh", "age.s"),
    thinSteps = 4000, # necessary 4000, takes 6.5 hours
    numSavedSteps = 100000, # 100000
    seed = 9382)
} else {
  file <- "chelsea_no_nas_res.Rdata"
  temp <- tempfile()
  download.file(paste0(path, file), destfile = temp, mode = "wb")
  con <- gzfile(temp, "rb")
  load(con)
  close(con)
  chelsea_no_nas_res <- coda::as.mcmc.list(chelsea_no_nas_res)
  chelsea_no_nas_res_diag <- diagnostic.summary(chelsea_no_nas_res)
  chelsea_no_nas_age_ranges <- 
    age.estim.summary(chelsea_no_nas_res_diag)
  threshold_no_nas <- chelsea_no_nas_res |> threshold.chains() |> 
    diagnostic.summary() |> threshold.matrix(mean_choice = "Mode")
}
```

Again, we only have a quick look at the worst quality measures:

```{r no nas diagnostics max min}
chelsea_no_nas_res_diag |> diagnostics.max.min()
```

Everything seems fine so we may proceed right to the correlation matrix:

```{r all methods nimble model, no NAs, correlation matrix}
options(knitr.kable.NA = '')
chelsea_no_nas_corr_mat <- corr.mat.mean(chelsea_no_nas_res) |> round(3)
chelsea_no_nas_corr_mat[lower.tri(chelsea_no_nas_corr_mat, diag = T)] <- NA
row.names(chelsea_no_nas_corr_mat) <- c("dental wear",  "pubic symphysis", 
                                 "auricular surface", "costochondral")
chelsea_no_nas_corr_mat[-4,-1] |> 
  knitr::kable (col.names = c("pubic symphysis", 
                                 "auricular surface", "costochondral"))
```

This time the correlation between pubic symphysis and auricular surface is nearly 0.4 and by far the highest between all pairs. The costochondral exhibits a mild positive correlation with pubic symphysis but a stronger negative correlation with dental wear. With the auricular surface, the costochondral is only weakly negatively correlated. Apart of the strong negative correlation with the costochondral, dental wear is only weakly positively correlated with the pubic symphysis and the auricular surface.

This is certainly not the place to explain these correlations in functional terms, especially considering the small sample number. Still, the difference in the pairwise correlations of pubic symphysis and auricular surface in the two runs merit attention. As shown below, the modal age of the individuals where all traits were assessable is the lowest of all sub-samples. This could mean that the correlation between pubic symphysis and auricular surface is stronger in younger age than in older age [on the difference between these two components see already @ref_21132].


### All traits, with conditional dependence

We are now ready to model all traits jointly together, including missing data.

```{r all methods jags model}
if ( runNewMCMC ) {
  chelsea_complete_nimble_res <- bay.ta(
    framework = "NIMBLE",
    method = chelsea_complete[,4:7],
    algorithm = "mnorm",
    eta = 1,
    minimum_age = 18,
    maximum_age = 100,
    parameters = c( "b", "a", "beta0", "beta", "thresh",  "age.s", "Ustar"),
    multicore = TRUE,
    thinSteps = 100, # necessary 5000, takes 10 hours
    numSavedSteps = 50000,
    seed = 3500)
} else {
  file <- "chelsea_complete_nimble_res.Rdata"
  temp <- tempfile()
  download.file(paste0(path, file), destfile = temp, mode = "wb")
  con <- gzfile(temp, "rb")
  load(con)
  close(con)
  chelsea_complete_nimble_res <- 
    coda::as.mcmc.list(chelsea_complete_nimble_res)
  chelsea_complete_nimble_diag <- 
    diagnostic.summary(chelsea_complete_nimble_res)
  chelsea_complete_nimble_age_ranges <- 
    age.estim.summary(chelsea_complete_nimble_diag)
  threshold_complete_nimble <- chelsea_complete_nimble_res |> 
    threshold.chains() |> diagnostic.summary() |> 
    threshold.matrix(mean_choice = "Mode")
}
```
  
  For the sake of completeness, we give the quality measures a quick glance:
  
```{r nimble complete diagnostics max min}
chelsea_complete_nimble_diag |> diagnostics.max.min()
```
  The values are satisfactory, so we inspect again the correlation matrix:
  
```{r all methods nimble model, correlation matrix}
options(knitr.kable.NA = '')
chelsea_complete_nimble_corr_mat <- 
  corr.mat.mean(chelsea_complete_nimble_res)
chelsea_complete_nimble_corr_mat[lower.tri(
  chelsea_complete_nimble_corr_mat, diag = T)] <- NA
row.names(chelsea_complete_nimble_corr_mat) <- 
  c("dental wear",  "pubic symphysis", "auricular surface", "costochondral")
chelsea_complete_nimble_corr_mat[-4,-1] |> 
  knitr::kable (digits = 3, col.names = c("pubic symphysis", 
                                 "auricular surface", "costochondral"))
```
  
  Here, again, as with the model limited to the auricular surface and the pubic symphysis, the correlation between these traits is very low. The highest positive correlation is now observed between pubic symphysis and dental wear. The costochondral is strongly negatively correlated with dental wear and especially the auricular surface. Again, the exploration of the possible functional reasons between these interesting observations is beyond the aims of this vignette.
  
  
### Threshold comparison

We are now in the position to inspect the results of all models so far. We will start with the thresholds which were extracted before with the helper function `threshold.matrix` from the MCMC results.

```{r thresholds table, echo = T}
options(knitr.kable.NA = '')
thresholds <- data.frame(rbind(
  c(threshold_dental, rep(NA, 4)),
  c(threshold_ps, rep(NA, 2)),
  threshold_as,
  threshold_costo,
  threshold_ps_as,
  threshold_no_nas,
  threshold_complete_nimble
) )
colnames(thresholds) <- c("I", "II", "III", "IV", "V", "VI", "VII")
rownames(thresholds) <- c("dental wear (dw)", "pubic symphysis (ps)", 
                          "auricular surface (as)", "costochondral (costo)", 
                          "ps_as - ps",  "ps_as - as", "no NAs - dw", 
                          "no NAs - ps", "no NAs - as", "no NAs - costo", 
                          "mnorm - dw", "mnorm - ps", "mnorm - as", 
                          "mnorm - costo")
thresholds |>  knitr::kable(digits = 0)
```

A few things merit attention. Firstly, especially with dental wear the runs with dental wear alone but also with the sample with all traits present, the result are not plausible: The first threshold is higher than the second. Even though this might partly be due to the chosen mean measure, the mode, and the result might look differently for arithmetic mean or median, it still points to an inherent problem of this particular trait. A further interesting aspects concern the threshold changes of the traits when they are estimated together with other traits. For example, for both pubic symphysis and auricular surface, the thresholds are higher for the model where they were analyzed together then when estimated on their own. At first sight perhaps rather counter-intuitively, higher threshold values will lead to lower age estimates.


### Age estimation comparison

Below is a summary of the age estimations of all model runs so far which was derived from the helper function `age.estim.summary()`. The arithmetic mean of the estimated parameters is chosen because only then the deterministic relationship between the Gompertz parameter $\beta$ (= `b` in the table below) and $\alpha$ and thus `M` holds. Only for the `highest density intervals`, also the median and the mode are shown.

```{r summary table, echo = F}
## Summary table
demo_sum <- data.frame(rbind(
  dental_wear = c(nrow(chelsea_dental), chelsea_dental_age_ranges[,1],
             chelsea_dental_age_ranges[5,2:3]),
  pubic_symph = c(nrow(chelsea_ps), chelsea_ps_age_ranges[,1],
             chelsea_ps_age_ranges[5,2:3]),
  auricular_surf = c(nrow(chelsea_as), chelsea_as_age_ranges[,1],
             chelsea_as_age_ranges[5,2:3]),
  costochondral = c(nrow(chelsea_costo), chelsea_costo_age_ranges[,1],
             chelsea_costo_age_ranges[5,2:3]),
  ps_as = c(nrow(chelsea_ps_as), chelsea_ps_as_age_ranges[,1],
             chelsea_ps_as_age_ranges[5,2:3]),
  no_NAs = c(nrow(chelsea_no_nas), chelsea_no_nas_age_ranges[,1],
             chelsea_no_nas_age_ranges[5,2:3]),
  complete_nimble =  c(nrow(chelsea_complete), 
                       chelsea_complete_nimble_age_ranges[,1],
             chelsea_complete_nimble_age_ranges[5,2:3])
) )
colnames(demo_sum) <- c("n", "b", "a", "M", "ø age", "HDI_mean", "HDI_median",
                        "HDI_mode")
demo_sum <- as.data.frame(do.call(cbind, lapply(demo_sum, unlist)))
demo_sum <- demo_sum |> mutate(across(c('b', 'a'), ~ round (., 3))) |>
  mutate(across(c('M','ø age', 'HDI_mean', 'HDI_median', 'HDI_mode'), 
                ~ round (., 1)))
demo_sum |> knitr::kable()
```

First of all, it is remarkable that for the first four rows which cover the single trait models, the results of all parameters are very close to each other. For the model with all traits and no `NAs`, a much lower modal age `M` is estimated. In contrast, for the pubic symphysis analysed together with the auricular surface as well as for the full model, a higher modal age is computed. Furthermore, for the single trait models, the averages of the `highest density intervals` range between 40 and 60 years. Especially the very high ranges for dental wear cast doubts on the usefulness of this age estimation method. In contrast, for the multivariate models the ranges are mostly below 30 years. For the multivariate models of the pubic symphysis with the auricular surface as well of all traits without `NAs`, these `HDIs` should cover the true age-at-death at the 0.95-credibility level, the default. Unfortunately, the number of cases with known age-at-death for these models is too low to test this proposition convincingly.

Interestingly, the full model exhibits about the same range in terms of `HDIs` as the model with pubis symphysis and auricular surface. The problem here is that because of the many missing entries, the correlation between traits cannot be taken fully into account, so we would expect that the true `HDI` is slightly higher.

As mentioned above, the full set includes 16 individuals with known age-at-death. For those, we can test if the computed `HDIs` include the known ages:

```{r all methods nimble model, binom test}
sequential.binom.test(chelsea_complete_nimble_res,
                      HDImass = 0.95,
                      known_age = chelsea_complete$known_age) |> 
  knitr::kable(digits = 3)
```

Of the 16 individuals, only 14 meet this condition instead of the at least expected 15. While the test is not significant due to the small sample size, this could imply that the `HDIs` are indeed not well-calibrated.


### All traits, without correlations (`JAGS`), but calibrated ages

Therefore, we will now add a calibration term to the model. We do this in the JAGS model because without the conditional dependence between traits this runs much faster.  The goal is to get to `HDI` ranges that are similar to the model with conditional dependence. This is achieved by adding a Gaussian distribution to the ages. It turns out that the similar `HDI` ranges are reached with an additional noise of 5 (years) in the parameter `error_sd`. The parameter to be monitored is changed from `age.s` to `age.s_c`.

```{r all methods jags model, with calibration}
if ( runNewMCMC ) {
  set.seed(634)
  chelsea_complete_cal_res <- bay.ta(
    framework = "JAGS",
    method = chelsea_complete[,4:7],
    minimum_age = 18,
    maximum_age = 100,
    error_sd = 5,
    parameters = c( "b", "a",  "beta0", "beta", "thresh",  "age.s_c"),
    multicore = TRUE,
    thinSteps = 5000, 
    numSavedSteps = 50000)
} else {
  file <- "chelsea_complete_cal_res.Rdata"
  temp <- tempfile()
  download.file(paste0(path, file), destfile = temp, mode = "wb")
  con <- gzfile(temp, "rb")
  load(con)
  close(con)
  chelsea_complete_cal_diag <- 
    diagnostic.summary(chelsea_complete_cal_res)
}
```

As always, a quick look at the summarized diagnostics gives us confidence to proceed:

```{r JAGS calibrated diagnostics  max min}
chelsea_complete_cal_diag |> diagnostics.max.min()
```

The following table illustrates the age estimate of the model with the calibration term.

```{r all methods jags model, with calibration, diagnostics}
chelsea_complete_cal_age_ranges <- 
  age.estim.summary(chelsea_complete_cal_diag, age_identifier = "age.s_c")
chelsea_complete_cal_age_ranges |> knitr::kable(digits = 3)
```

The age ranges are now around 30 years. The following gallery of plots demonstrate the relation between the mode of each age estimate and the range of the `HDIs`.

```{r all major model, hdi ranges, fig.width=7, fig.height=6}
chelsea_no_nas_res_diag_red_ages <-
  chelsea_no_nas_res_diag[grep("^age.s",
                                 rownames(chelsea_no_nas_res_diag)),]
chelsea_ps_as_res_diag_red_ages <-
  chelsea_ps_as_res_diag[grep("^age.s",
                                 rownames(chelsea_ps_as_res_diag)),]
chelsea_complete_nimble_diag_red_ages <-
  chelsea_complete_nimble_diag[grep("^age.s",
                                 rownames(chelsea_complete_nimble_diag)),]
chelsea_complete_cal_diag_red_ages <-
  chelsea_complete_cal_diag[grep("^age.s_c",
                                 rownames(chelsea_complete_cal_diag)),]
ggpubr::ggarrange(
   ggplot(chelsea_ps_as_res_diag_red_ages,
         aes(y = HDIhigh - HDIlow, x = Mode)) + geom_point() +
    ylim(0,60) + xlim(15,100) + 
    ggtitle("Pubic symphysis & auricular surface") + 
    theme_light(),
    ggplot(chelsea_no_nas_res_diag_red_ages,
         aes(y = HDIhigh - HDIlow, x = Mode)) + geom_point() +
    ylim(0,60) + xlim(15,100) + ggtitle("no NAs") + theme_light(),
  ggplot(chelsea_complete_nimble_diag_red_ages,
         aes(y = HDIhigh - HDIlow, x = Mode)) + geom_point() +
    ylim(0,60) + xlim(15,100) + 
    ggtitle("complete set with NIMBLE, uncalibrated") + 
    theme_light(),
  ggplot(chelsea_complete_cal_diag_red_ages,
         aes(y = HDIhigh - HDIlow, x = Mode)) + geom_point() +
    ylim(0,60) + xlim(15,100) + ggtitle("complete set with JAGS, calibrated") + 
    theme_light(),
   ncol = 2, nrow = 2
)
```

As noted previously (§§), the HDI ranges are lowest in early life and increase from there onwards. However, at about an age mode of 60--70 years, they start to shrink again.

The plots of the joined analysis of the auricular surface and the pubic symphysis as well as that for the data set without `NAs` look very similar, despite the fact that the first model contained much more entries than the second. In both cases, the point estimates form a rather clean arc. In contrast, the complete model looks rather messy with some extreme outliers, presumably from individuals were only the weakest method, dental wear, was assessable. Furthermore, the clean arc of the first two models has disappeared because middle-aged individuals were assessed overall younger than in the first two models. In comparison, the complete model with the calibration term of 5 years has again moved closer to the model without `NAs`. This we would take as a hint that the estimates of the model with calibration term are more conservative.

Next, we compare the calibrated estimated ages-at-death with the true ages-at-death known for 16 individuals. 

```{r all methods jags model, with calibration, simple comparison with known age-at-death}
chelsea_complete$estimated_age <- 
  round(chelsea_complete_cal_diag_red_ages[,5],1)
chelsea_complete[,c("AGE", "known_age", "estimated_age")] |> 
  na.omit() |> knitr::kable()
```

The agreement between known and estimated age-at-death is already very satisfying. Eye-balling shows only three more severe outliers (nos. 24, 114 and 161). The traditional aging method is wrong in three cases (nos. 62, 114, 161). For a more systematic comparison, we evoke the helper function `age.comp.summary()`.

```{r all methods jags model, known age-at_death}
summary_list <- lapply(c("Mode", "Median", "Mean"), function(choice) {
  age.comp.summary(mcmc_list = chelsea_complete_cal_res, 
                   known_age = chelsea_complete$known_age,
            age_identifier = "age.s_c",
                   mean_choice = choice)})
summary_mat <- do.call(rbind, summary_list)
rownames(summary_mat) <- c("Mode", "Median", "Mean")
summary_mat |> t() |> knitr::kable(digits = 2)
```

Above the resulting coverage for 95% credibility was computed. As already pointed out above, the first six frequentist quality measures are dependent on the point estimate and therefore malleable to the chosen measure of the mean. The mode is best in terms of the first four quality measures while the arithmetic mean performs better for `Inaccuracy` and `RMSE`. The median is somehow in-between. The two Bayesian measures `TMNLP` and `CRPS`, in contrast, do not change, regardless which point estimate was chosen.

When reviewing the quality measures itself (see §§ for a short review of some tests of age estimation methods), the performance is quite good. Most single methods will not reach an `Inaccuracy` of 7.3 years or a `CRPS` of 5.25.

A formal test of the coverage is the binomial test which checks if the resulting numbers deviate significantly from expectation. Below, we do this for several credibility ranges. Due to the small sample size of only 16 age-known individuals, we cannot expect any significant result.

```{r all methods jags model, binom test}
sequential.binom.test(chelsea_complete_cal_res,
                      HDImass = c(seq(0.5, 0.9, 0.1), 0.95),
                      age_identifier = "age.s_c",
                      known_age = chelsea_complete$known_age) |> 
  knitr::kable(digits = 3)
```

We show the relationship between known and estimated age-at-death also in a series of plots which are generated with the helper function `age.comp.plot()`.

```{r all methods jags model, bay.ta.plot, fig.width=7, fig.height=6}
age.comp.plot(x = chelsea_complete_cal_diag,
            age_identifier = "age.s_c",
            known_age = chelsea_complete$known_age,
            mean_choice = "Mode")
```

Only one `HDI` range does not include the true age (left top). The top right plot shows the Gompertz function based on the estimates of all individuals (in black) with the density of the known ages-at-death (in red). That the latter represent a part of the population seems from this plot at least not implausible. The two bottom plots illustrate the differences of the point estimate from the known age-at-death. Especially the right bottom plot also shows the systematic deviation which leads to a *residual slope* of 0.21. Most differences between estimated and known age are at about 10 years or below but one case deviates by more than 20 year.


## Conclusions

One of the main aims of the above workflow was to show how the *highest density intervals* can be calibrated when several traits are observed but the data is incomplete. This entails to run a multivariate normal model with only those cases that do not contain any `NA` values and then set the calibration in such a way that the ensuing `HDIs` emulate that from the model with no `NAs`. For the Chelsea 'Old church' data set, this had lead to age ranges that fit to the known ages-at-death.

Apart from this, two further aspects which become apparent with the Chelsea 'Old church' data set merit attention. The first concerns the different mean and modal ages for the different model based on different traits and trait combinations. It has to be kept in mind that the ensuing samples are not identical and only partly overlap. That means that the different mean ages might well reflect true differences in the age composition of the respective samples. Especially in the case of the model of no `NAs`, that is the one where all traits were assessable, the low mean age is noteworthy, despite the fact that the plot with the ages and the `HDIs` looked very similar to the one with auricular surface and pubic symphsis. Therefore, we have to consider that age-at-death and completeness of trait observations is very probably not independent from one another. This suggests that missingness is not random with respect to age, which may bias multivariate estimates.
 
 The second aspects relates to the *highest density intervals*: For the multivariate normal models, these were around 30 years of age. This is the same range which was found for the DRNNAGE system (§§) but also for the new ABDOU transition analysis (manuscript in prep.). It seems quite possible that with methods based on pathological changes in the skeleton, as those age markers essentially are [@FUCHS20261], we cannot get below this threshold of 30 years. This seems to equal an inaccuracy level of about 6--7 years or a `CRPS` of about 4--5 years.
 
---

## References
