---
title: "Grouping of data"
bibliography: ../REFERENCES.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Load libraries

```{r setup, message = F}
library(baytaAAR)
library(ggpubr)
library(ggridges)
library(tidybayes)
library(ggplot2)
library(tidyr)
library(dplyr)
```


## Introduction

While `bay.ta()` estimates age at the individual level, many research questions operate at the group level (e.g., comparing burial populations or social categories). Because `bay.ta()` does not (yet) contain a grouping variable as would be the case in a hierarchic model, the following approach aggregates posterior age distributions *post hoc* without modifying the underlying model. In this way, we can compare sub-groups within one data set. This purpose serves the helper function `prob.cat()`. We demonstrate its functioning here with the two variables `AGE` and `SEX` of the osteological analysis of Chelsea 'Old church'. More meaningful choices could be sub-populations according to location, burial customs or grave goods. However, the principle would be the same.


## Downloading the data

We re-use here the data already used for the vignette on Chelsea 'Old church' (`vignette("worked_example")`). So, the respective MCMC data is downloaded from a `Github` repository.

```{r paths for rda-files}
path <- "https://raw.githubusercontent.com/ISAAKiel/Chelsea_mcmc/main/"
file <- "chelsea_complete_cal_res.Rdata"
temp <- tempfile()
if (Sys.info()[["sysname"]] == "Darwin") {
  download.file(paste0(path, file), destfile = temp, method = "curl", mode = "wb")
} else {
  download.file(paste0(path, file), destfile = temp, method = "libcurl", mode = "wb")
}
con <- gzfile(temp, "rb")
load(con)
close(con)
```

```{r loading of data, echo=F, warning=FALSE}
temp <- tempfile()
download.file(
  "https://dams.londonmuseum.org.uk/m/2dc89b01e62759ff/original/PMCOC.zip",
  temp, mode = "wb")
unzip(temp, "PMCOC_age_estimates.lst")
chelsea_age <- read.delim("PMCOC_age_estimates.lst",
                          sep = "|",
                          header = TRUE,
                          stringsAsFactors = FALSE,
                          row.names = NULL,
                          fill = TRUE,
                          strip.white = TRUE)

# absolute ages according to Cowie et al. 2008, 41 table 6
chelsea_known_age <- data.frame(
  CONTEXT = c(35, 147, 198, 462, 525, 622, 654, 681, 701, 713,
              722, 744, 792, 976, 980, 990, 1133),
  known_age = c(60, 67, 44, 61, 70, 84, 66, 84, 78, 68, 56, 70,
                68, 82, 54, 32, 70) )

chelsea_age_col <- colnames(chelsea_age)[-7]
chelsea_age <- chelsea_age[,-14]
colnames(chelsea_age) <- chelsea_age_col

chelsea_adults <- 
  chelsea_age |> dplyr::filter(grepl("^(ADULT|UNCLASSIFIED AD)", AGE))

chelsea_adults_wide <- chelsea_adults |>
  dplyr::mutate(VALUE = as.numeric(VALUE)) |>
  tidyr::pivot_wider(
    id_cols = c(CEMETERY, SITECODE, PERIOD, LU_INT, E_DATE, L_DATE,
                CONTEXT, SEX, AGE, TRAIT_TYPE),
    names_from = EXPANSION,
    values_from = VALUE
  )

# combine both tables
chelsea <- merge(chelsea_adults_wide, chelsea_known_age,
                     by = "CONTEXT",
                     all.x = TRUE)

# limit data set to relevant columns
chelsea_mat <- chelsea[,c(1, 8, 9, 12,13,14,15,18)]

chelsea_mat[chelsea_mat == 9] <- NA
chelsea_complete <- chelsea_mat[rowSums(is.na(chelsea_mat[,4:7])) != 4, ]
chelsea_complete$AGE <- factor(chelsea_complete$AGE,
                               levels = c("ADULT 18-25 YEARS",
                                          "ADULT 26-35 YEARS",
                                          "ADULT 36-45 YEARS",
                                          "ADULT >46 YEARS",
                                          "UNCLASSIFIED ADULT"),
                               ordered = TRUE)
chelsea_complete$SEX <- factor(chelsea_complete$SEX,
                          levels = c("FEMALE",
                                     "FEMALE?",
                                     "INTERMEDIATE",
                                     "MALE?",
                                     "MALE"),
                          ordered = TRUE)
```


## Probability densities in practice

### Data input to `prob.cat()`

Apart from the `coda::mcmc.list()` resulting from `bay.ta()`, `prob.cat()` expects a character string (`age_identifier`) specifying the row name of the age estimates (either `age.s` or `age.s_c` as here), a vector of the grouping variable (`group_vec`) which needs to be of the same length as the number of age estimates and a specification of the mode switch. The `summed` option produces a pooled (mixture) distribution of all individuals within a group, reflecting the combined posterior probability mass. In contrast, `mean` averages the posterior values across individuals per iteration, yielding a distribution of group-level central tendencies. The results are visualized with `ggplot2` [@wickham_2016] and the additional packages `tidybayes` [@kay_2024], `ggpubr` [@kassambara_2025], and `ggridges` [@wilke_2025].

For the blue vertical lines in the left plots below, we need a column for the estimated ages. Again, we recycle here code from the vignette on Chelsea that adds this column to the original data.

```{r comparison with known age-at-death}
chelsea_complete_cal_diag <- 
    diagnostic.summary(chelsea_complete_cal_res)
chelsea_complete_cal_diag_red_ages <-
  chelsea_complete_cal_diag[grep("^age.s_c",
                                 rownames(chelsea_complete_cal_diag)),]
chelsea_complete$estimated_age <- 
  round(chelsea_complete_cal_diag_red_ages[,5],1)
```


### Chelsea 'Old church': Age

As first example, we group the posterior probability densities by the osteological age categories as determined by the team of the Museum of London. Since both approaches rely on the same underlying traits, a broad agreement is expected, although systematic differences may arise from the probabilistic modelling and calibration. 

```{r all methods jags model, probability densities AGE, fig.width=7, fig.height=5}
chelsea_summed_prob_age <- prob.cat(mcmc_list = chelsea_complete_cal_res,
                                  age_identifier = "age.s_c",
                                  group_vec = chelsea_complete$AGE,
                                  mode = "summed")
chelsea_mean_prob_age <- prob.cat(mcmc_list = chelsea_complete_cal_res,
                                    age_identifier = "age.s_c",
                                  group_vec = chelsea_complete$AGE,
                                    mode = "mean")
ggpubr::ggarrange(
  ggplot(chelsea_summed_prob_age, aes(x = x, y = category, height = y)) +
    ggridges::geom_density_ridges(
      stat = "identity",
      colour = "darkgrey",
      fill = "lightgrey",
      alpha = 0.7,
      scale = 0.9,
      rel_min_height = 0.01
    ) +
    scale_x_continuous(limits = c(18, 100), breaks = seq(20, 100, by = 10)) +
    xlab("Probability density of estimated age") +
    ylab("Age category") +
    scale_y_discrete(expand = expansion(mult = c(0.05, 0.3))) +
    ggridges::theme_ridges(font_size = 11) +
    geom_point(
      data = chelsea_complete,
      aes(x = estimated_age, y = as.numeric(AGE) + 0.075),
      color = "blue", shape = '|', size = 4, alpha = 0.8,
      position = position_jitter(height = 0), inherit.aes = FALSE
    ),
  ggplot(chelsea_mean_prob_age, aes(x = value, y = category)) +
    tidybayes::stat_halfeye(
      .width = 0.95,
      adjust = 0.6,
      slab_size = 0.6,
      point_interval = tidybayes::mode_hdi,
      slab_colour = "darkgrey",
      fill = "lightgrey",
      linewidth = 0.8
    ) +
    scale_x_continuous(limits = c(18, 100), breaks = seq(20, 100, by = 10)) +
    xlab("Posterior mean age") +
    ylab("Age category") +
    scale_y_discrete(expand = expansion(mult = c(0.05, 0.3))) +
    ggridges::theme_ridges(font_size = 11)
)

```

The left plot displays summed probability densities of all estimates of one osteological age group combined with the mode of the individual estimates as blue short vertical lines. The densities are normalized within each group and therefore do not reflect group size. The plot somehow resembles similar plots of calibrated C14 data. The overall agreement between the osteological age estimates and the Bayesian estimates is very good. However, a close inspection shows that especially the age group 36--45 years according to the Bayesian estimates extends far beyond the proposed limit of 45 years. This suggests that the traditional age category may truncate older individuals whose trait expressions overlap with higher age ranges. In the right plot the mean of this age group is fixed at about 50 years. The means of the other age groups are within the limits of the osteological age ranges, though at the upper range.


### Chelsea 'Old church': Sex

The second example groups the probability densities according to the sex determinations.

```{r all methods jags model, probability densities SEX, fig.width=7, fig.height=5}
chelsea_summed_prob_sex <- prob.cat(mcmc_list = chelsea_complete_cal_res,
                                  age_identifier = "age.s_c",
                                  group_vec = chelsea_complete$SEX,
                                  mode = "summed")
chelsea_mean_prob_sex <- prob.cat(mcmc_list = chelsea_complete_cal_res,
                                    age_identifier = "age.s_c",
                                    group_vec = chelsea_complete$SEX,
                                    mode = "mean")

ggpubr::ggarrange(
  ggplot(chelsea_summed_prob_sex, aes(x = x, y = category, height = y)) +
    ggridges::geom_density_ridges(
      stat = "identity",
      colour = "darkgrey",
      fill = "lightgrey",
      alpha = 0.7,
      scale = 0.9,
      rel_min_height = 0.01
    ) +
    scale_x_continuous(limits = c(18, 100), breaks = seq(20, 100, by = 10)) +
    xlab("Probability density of estimated age") +
    ylab("Sex category") +
    scale_y_discrete(expand = expansion(mult = c(0.05, 0.3))) +
    ggridges::theme_ridges(font_size = 11) +
    geom_point(
      data = chelsea_complete,
      aes(x = estimated_age, y = as.numeric(SEX) + 0.075),
      color = "blue", shape = '|', size = 4, alpha = 0.8,
      position = position_jitter(height = 0), inherit.aes = FALSE
    ),

  ggplot(chelsea_mean_prob_sex, aes(x = value, y = category)) +
    tidybayes::stat_halfeye(
      .width = 0.95,
      adjust = 0.6,
      slab_size = 0.6,
      point_interval = tidybayes::mode_hdi,
      slab_colour = "darkgrey",
      fill = "lightgrey",
      linewidth = 0.8
    ) +
    scale_x_continuous(limits = c(18, 100), breaks = seq(20, 100, by = 10)) +
    xlab("Posterior mean age") +
    ylab("Sex category") +
    scale_y_discrete(expand = expansion(mult = c(0.05, 0.3))) +
    ggridges::theme_ridges(font_size = 11)
)
```

Compared to `AGE`, the `SEX` groups display no clear pattern. Their means on the right plot are all around 55 years, and their *highest density intervals*, the black line at the bottom of the densities extending to the left and right from the black circle, overlap to a large extent. The lack of systematic differences between sex categories suggests that the estimated age distributions are largely independent of osteological sex classification in this data set. Nevertheless, the bimodal distribution of the females raises some questions. It seems possible that the peak in their 20s reflects heightened maternal mortality or that access to the crypt favored either very young or comparatively old women.

Overall, the examples above illustrate how posterior age distributions can be flexibly aggregated to explore group-level patterns and assess how posterior age distributions vary across categorical variables without modifying the underlying model.

---

## References
